---
title: "Classification"
author: "Nate Wells"
date: "11/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

The plot below represents the predictor space (on $X_1$ and $X_2$) with a training data set plotted and the class of their response variable indicated by shape and color.

```{r}
set.seed(100)
n <- 6
circles <- data.frame(
  shape = "circle",
  x1 = runif(n+1, 0.05,.95),
  x2 = runif(n+1, .25,.95 )
)
triangles <- data.frame(
  shape = "triangle",
  x1 = runif(n, 0.05,.6),
  x2 = runif(n,  0.05,.6)
)
squares <- data.frame(
  shape = "square",
  x1 = runif(n-2, 0.0,.95),
  x2 = runif(n-2, 0.1,.75)
)
shapes <- rbind(circles, triangles, squares) %>% mutate(shape = as.factor(shape))


g<- ggplot(shapes, aes(x = x1, y = x2, col = shape, shape = shape)) +
  geom_point(size = 4) +
  scale_x_continuous(expand = c(0, 0) , limits = c(0, 1)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1)) +
  theme_bw()

g
```

1. If we consider this a classification tree without any splits yet, what would be the prediction for every test observation?

2. What is the (training) misclassificaiton rate?

3. What is the Gini Index for this node?

4. What is the information (entropy)?

5. Add a straight line to one of your axes that splits the predictor space into two regions. Choose the split in a ay that you think will lead to the best overall improvement in metrics listed above. Label the new regions R1 and R2 and calculate the metrics for each:

a. For each region, what is the predicted class?

b. For each region, what is the misclassification rate?

c. For each region, what is the Gini index?

d. For each region, what is the information (entropy)?

6. Draw the very simple regression tree corresponding to the split you made.